Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-20 paper.qmd
Start updating the discussion

The models are fitted to these layouts. @fig-one-dens_clust-error b1 and b2 show the fitted models in projections of the $7\text{-}D$ space. In these projections, the differences in the distribution of high-dimensional model errors can be observed. The uniform C-shaped cluster displays randomly distributed model errors, while the dense C-shaped cluster shows high model errors at the sparse end. In the dense C-shaped cluster, the high model errors occur because there are fewer data points positioned within the bins at the sparse end of the structure.

@@

With the model fit, some quirks can be observed. The simulated $4\text{-}D$ data having five Gaussian clusters is used. tSNE, and UMAP which showed the fitted model as *crumpled \gD{} sheet* that stretches in all four dimensions evident of preserving local structure. The flat model generated by PaCMAP evident the fail to capture the within cluster variation.

@@

The **digit: 1** example further illustrates the model's ability to accurately capture non-linear structures and provide additional information. Key findings include a twisted pattern that compresses the structure in some projections and long line segments that detect anomalies.

@@

To extend layouts beyond \kD{}, when \gD{} is clearly inadequate, can use k-means instead of bin centroids.

>>>>

The models are fitted to these layouts. @fig-one-dens_clust-error b1 and b2 show the fitted models in projections of the $7\text{-}D$ space. In these projections, the differences in the distribution of \pD{} model errors can be observed. The uniform C-shaped structure displays randomly distributed model errors, while the dense C-shaped structure shows high model errors at the sparse end. In the dense C-shaped structure, the high model errors occur because there are fewer data points positioned within the bins at the sparse end of the structure.

@@

With the model fit, several observations can be made. The simulated $4\text{-}D$ data, which contains five Gaussian clusters, was analyzed. Both tSNE and UMAP demonstrated the fitted model as a *crumpled \gD{} sheet* that stretches across all four dimensions, highlighting its ability to preserve local structure. In contrast, the flat representation generated by PaCMAP indicates its failure to capture the variation within the clusters. This highlights how different NLDR methods organize the data points differently.

Another usability of our approach is how NLDR methods behave with different density of the data structure. The tSNE layout of simulated $7\text{-}D$ C-shaped data with different densities are used. The high \pD{} model errors are scattered to a corner of the \gD{} layout of dense C-shaped structure. In the dense C-shaped structure, the high model errors occur because there are fewer data points positioned within the bins at the sparse end of the structure.

@@

The digit 1 of **Hand-written digits** data shows how the model accurately capture the non-linear structure. Also, twisted pattern of the model evident that

example further illustrates the model's ability to accurately capture non-linear structures and provide additional information. Key findings include a twisted pattern that compresses the structure in some projections and long line segments that detect anomalies.

@@

Furthermore, to extend layouts beyond \kD{}, when \gD{} is clearly inadequate, k-means can be used as an alternative to bin centroids.
